Parallel algorithm for solving multicriterial optimization problems using elements of machine learning

### 2025-04-06 2025/RSCD_MCO/RSD-2025.tex
fix MCO HW->HV

As a criterion for the quality of the found Pareto set, the indicator HW \cite{Evtushenko2014,pymoo,AGP_ML} was applied. Higher value of the indicator corresponds to better solution. To assess the performance of the algorithm, the number of trials carried out when solving the problem are given. However, when solving complex optimization problems, one trial can take a significant time. In such conditions, the reduction of the number of trials reduces the total computation time in general. In the presented experimental results, the number of trials spent on solving each problem and the HW indicators were averaged according to the number of solved problems.

@@

Fig. \ref{fig5} shows the dependence of the average value of the HW index on the average number of trials for the sequential algorithm. Several solution accuracy values $\varepsilon = \{0.1,0.05,0.01\}$ were fixed for plotting the graph for algorithm (\ref{eq:13}). 100 problems were solved for each accuracy value and the average results were plotted on a graph. The position of the resulting curve for the chosen algorithm to the left and above corresponds to the best behavior of the algorithm as a whole. During the experiments the parameter $\alpha$ from (15) varied, Fig. \ref{fig5} shows only the best results for each of the variants.

\begin{figure}
\center
\includegraphics[width=0.8\textwidth]{fig5.png}
\caption{Comparison of HW index values for sequential algorithm implementation.} \label{fig5}
\end{figure}

Based on the results of experiments with sequential implementation, all approaches using machine learning models exceed the MGSA algorithm by more than two times. Modifications of the algorithm can be distinguished that demonstrate the best results both in terms of the average number of iterations and in terms of the average HW. These include approaches using SVC (Linear, Rbf cores), a distance calculation approach, and neural networks with two hidden layers. Modification with one hidden layer shows slightly worse results. Modification  MGSA Poly SVC with polynomial core showed least improvement over MGSA. It is also worth noting that the modification of the MGSA Rbf SVC algorithm in the case of low accuracy demonstrates the best quality.

@@

According to Table \ref{tab:1}, MGSA Dist, MGSA Linear SVC and MGSA NN algorithms with two hidden layers demonstrate the best results. To analyze the quality of the resulting solution, the values   of the HW index are given in Table \ref{tab:2}. The table also shows three best modifications.

\begin{table}[ht]
\centering
\caption{Average HW Index in dependence on number of processes}
\label{tab:2}
\resizebox{0.8\textwidth}{!}{%
\begin{tabular}{ccccccc}
\hline
Number of processes                   & \textbf{1}      & \textbf{2}      & \textbf{4}      & \textbf{8}      & \textbf{16}     & \textbf{20}     \\ \hline
\textbf{MGSA}                         & \textbf{114.91} & \textbf{114.79} & \textbf{114.84} & \textbf{114.90} & \textbf{115.00} & \textbf{114.96} \\
MGSA Linear SVC, $\alpha$ = 0.05           & 114.44          & 114.65          & 114.72          & 114.74          & 114.91          & 114.96          \\
\textbf{MGSA   Poly SVC, $\alpha$ = 0.05} & \textbf{114.61} & \textbf{114.64} & \textbf{114.71} & \textbf{114.8}  & \textbf{114.95} & \textbf{115.01} \\
MGSA RBF SVC, $\alpha$ = 0.04           & 114.54          & 114.54          & 114.73          & 114.89          & 114.92          & 114.96          \\
MGSA Dist, $\alpha$ = 0.01                 & 114.66          & 114.55          & 114.69          & 114.73          & 114.92          & 114.93          \\
MGSA NN 1 layer, $\alpha$ = 0.05           & 114.77          & 114.61          & 114.72          & 114.81          & 114.96          & 114.98          \\
\textbf{MGSA NN 2 layers, $\alpha$ = 0.03} & \textbf{114.74} & \textbf{114.54} & \textbf{114.71} & \textbf{114.76} & \textbf{114.98} & \textbf{115.01} \\ \hline
\end{tabular}%
}
\end{table}

@@

\begin{figure}
\center
\includegraphics[width=0.8\textwidth]{fig6.png}
\caption{Dependence of the mean value of the index HW from the trial number per process for the case of 20 processes.} \label{fig6}
\end{figure}

Fig. \ref{fig6} demonstrates that as the number of processes used increases, all algorithms are significantly ahead of the MGSA method, which does not use machine learning models, in both the average number of trials per process and the HW indicator, especially if high solution accuracy is to be ensured. All modifications of the algorithm except MGSA Poly SVC and MGSA NN with one layer show similar quality. A simpler modification MGSA NN with one hidden layer with a small number of trials lags behind in quality and aligns with the rest when increasing the number of trials. This may indicate that this model can be useful when high-precision calculations are needed, in other cases it is recommended to use other machine learning models. At the same time, a simple model can learn much faster.

Fig. \ref{fig7} shows the value of the HW index depending on the number of trials when the number of used processes changes. Since MGSA NN 2 layers demonstrates better results than MGSA NN 1 layer, the information about the last modification are not given. According to Fig. \ref{fig7}, all presented models show a significant increase in the HW index and a decrease in the trial number per process when increasing the number of processes. Approach without using machine learning methods demonstrate similar dynamics when the target accuracy is high, however, at lower values, changes in the number of counted criteria on one process become less noticeable. It is also worth noting that, unlike the basic MGSA algorithm, all modifications with machine learning elements do not lose quality when switching to a parallel implementation.

\begin{figure}
\center
\includegraphics[width=0.8\textwidth]{fig7.png}
\caption{Dependence of the HW index on the number of trials executed for different number of processes.} \label{fig7}
\end{figure}

>>>>

As a criterion for the quality of the found Pareto set, the indicator HV \cite{Evtushenko2014,pymoo,AGP_ML} was applied. Higher value of the indicator corresponds to better solution. To assess the performance of the algorithm, the number of trials carried out when solving the problem are given. However, when solving complex optimization problems, one trial can take a significant time. In such conditions, the reduction of the number of trials reduces the total computation time in general. In the presented experimental results, the number of trials spent on solving each problem and the HV indicators were averaged according to the number of solved problems.

@@

Fig. \ref{fig5} shows the dependence of the average value of the HV index on the average number of trials for the sequential algorithm. Several solution accuracy values $\varepsilon = \{0.1,0.05,0.01\}$ were fixed for plotting the graph for algorithm (\ref{eq:13}). 100 problems were solved for each accuracy value and the average results were plotted on a graph. The position of the resulting curve for the chosen algorithm to the left and above corresponds to the best behavior of the algorithm as a whole. During the experiments the parameter $\alpha$ from (15) varied, Fig. \ref{fig5} shows only the best results for each of the variants.

\begin{figure}
\center
\includegraphics[width=0.8\textwidth]{fig5.png}
\caption{Comparison of HV index values for sequential algorithm implementation.} \label{fig5}
\end{figure}

Based on the results of experiments with sequential implementation, all approaches using machine learning models exceed the MGSA algorithm by more than two times. Modifications of the algorithm can be distinguished that demonstrate the best results both in terms of the average number of iterations and in terms of the average HV. These include approaches using SVC (Linear, Rbf cores), a distance calculation approach, and neural networks with two hidden layers. Modification with one hidden layer shows slightly worse results. Modification  MGSA Poly SVC with polynomial core showed least improvement over MGSA. It is also worth noting that the modification of the MGSA Rbf SVC algorithm in the case of low accuracy demonstrates the best quality.

@@

According to Table \ref{tab:1}, MGSA Dist, MGSA Linear SVC and MGSA NN algorithms with two hidden layers demonstrate the best results. To analyze the quality of the resulting solution, the values   of the HV index are given in Table \ref{tab:2}. The table also shows three best modifications.

\begin{table}[ht]
\centering
\caption{Average HV Index in dependence on number of processes}
\label{tab:2}
\resizebox{0.8\textwidth}{!}{%
\begin{tabular}{ccccccc}
\hline
Number of processes                   & \textbf{1}      & \textbf{2}      & \textbf{4}      & \textbf{8}      & \textbf{16}     & \textbf{20}     \\ \hline
\textbf{MGSA}                         & \textbf{114.91} & \textbf{114.79} & \textbf{114.84} & \textbf{114.90} & \textbf{115.00} & \textbf{114.96} \\
MGSA Linear SVC, $\alpha$ = 0.05           & 114.44          & 114.65          & 114.72          & 114.74          & 114.91          & 114.96          \\
\textbf{MGSA   Poly SVC, $\alpha$ = 0.05} & \textbf{114.61} & \textbf{114.64} & \textbf{114.71} & \textbf{114.8}  & \textbf{114.95} & \textbf{115.01} \\
MGSA RBF SVC, $\alpha$ = 0.04           & 114.54          & 114.54          & 114.73          & 114.89          & 114.92          & 114.96          \\
MGSA Dist, $\alpha$ = 0.01                 & 114.66          & 114.55          & 114.69          & 114.73          & 114.92          & 114.93          \\
MGSA NN 1 layer, $\alpha$ = 0.05           & 114.77          & 114.61          & 114.72          & 114.81          & 114.96          & 114.98          \\
\textbf{MGSA NN 2 layers, $\alpha$ = 0.03} & \textbf{114.74} & \textbf{114.54} & \textbf{114.71} & \textbf{114.76} & \textbf{114.98} & \textbf{115.01} \\ \hline
\end{tabular}%
}
\end{table}

@@

\begin{figure}
\center
\includegraphics[width=0.8\textwidth]{fig6.png}
\caption{Dependence of the mean value of the index HV from the trial number per process for the case of 20 processes.} \label{fig6}
\end{figure}

Fig. \ref{fig6} demonstrates that as the number of processes used increases, all algorithms are significantly ahead of the MGSA method, which does not use machine learning models, in both the average number of trials per process and the HV indicator, especially if high solution accuracy is to be ensured. All modifications of the algorithm except MGSA Poly SVC and MGSA NN with one layer show similar quality. A simpler modification MGSA NN with one hidden layer with a small number of trials lags behind in quality and aligns with the rest when increasing the number of trials. This may indicate that this model can be useful when high-precision calculations are needed, in other cases it is recommended to use other machine learning models. At the same time, a simple model can learn much faster.

Fig. \ref{fig7} shows the value of the HV index depending on the number of trials when the number of used processes changes. Since MGSA NN 2 layers demonstrates better results than MGSA NN 1 layer, the information about the last modification are not given.

According to Fig. \ref{fig7}, all presented models show a significant increase in the HV index and a decrease in the trial number per process when increasing the number of processes. Approach without using machine learning methods demonstrate similar dynamics when the target accuracy is high, however, at lower values, changes in the number of counted criteria on one process become less noticeable. It is also worth noting that, unlike the basic MGSA algorithm, all modifications with machine learning elements do not lose quality when switching to a parallel implementation.

\begin{figure}
\center
\includegraphics[width=0.8\textwidth]{fig7.png}
\caption{Dependence of the HV index on the number of trials executed for different number of processes.} \label{fig7}
\end{figure}
